\phantomsection
\addcontentsline{toc}{chapter}{\bf ABSTRACT}

\begin{center}
    \large{\bf ABSTRACT}
\end{center}

This project addresses the problem of identifying optimal conversion paths in large-scale customer journey graphs. User interaction sequences are modelled as a directed weighted graph $G=(V,E)$, where edge weights encode the negative log-probability of each state transition, reducing the maximum-probability path search to shortest-path form amenable to Dijkstra's algorithm. Existing acceleration techniques---A* search, beam search, contraction hierarchies---either demand a problem-specific admissible heuristic, sacrifice optimality guarantees, or require heavy preprocessing ill-suited to behavioural data.

Two algorithms are proposed and evaluated. The \textbf{Baseline Dijkstra} algorithm provides globally optimal solutions in $O(|E|\log|V|)$ time. The \textbf{Probability-Pruned Dijkstra} extends the baseline with a threshold parameter $\tau$ that discards low-probability partial paths during traversal, reducing the effective search space to $O(|E'|\log|V|)$ with $|E'|\leq|E|$ while provably converging to the baseline as $\tau\to 0$.

Both algorithms were evaluated on synthetic graphs up to 10,000 nodes (2,160 experiment rows) and validated on two real-world e-commerce clickstream datasets: RetailRocket (2.7M events) and RecSys~2015 YOOCHOOSE (33M clicks), totalling 840 real-data rows across fixed and adaptive $\tau$ regimes. On synthetic benchmarks, wall-clock speedups range from $3.3\times$ ($\tau=0.001$) to $738\times$ ($\tau=0.5$). On real-world item-level graphs under fixed $\tau$, speedups reach $18{,}882\times$ (RetailRocket) and $2{,}309\times$ (RecSys~2015). An adaptive-$\tau$ regime---where the threshold is calibrated to each query pair's baseline path probability---raises real-data path-found rates from near 0\% to over 80\% while preserving optimality. Crucially, the optimality gap is exactly $0.0000\%$ across all 706 path-found rows---the pruned algorithm either returns the globally optimal path or prunes it entirely, confirming a formal ``all-or-nothing'' property. These results suggest that threshold-based pruning offers a practical, parameter-controlled acceleration strategy for probabilistic journey optimization on graphs with tens of thousands of nodes, without sacrificing solution quality when paths are found.
