\chapter{ALGORITHM DESIGN, DATA STRUCTURES, AND IMPLEMENTATION}

\section{System Overview}

The \textit{Customer Journey Path Optimization} system models user interactions as a directed weighted graph $G = (V, E)$, where each node $v \in V$ represents a discrete interaction state (e.g., \textit{Landing Page}, \textit{Search}, \textit{Product Page}, \textit{Cart}, \textit{Checkout}), and each directed edge $(u,v) \in E$ represents a user transition between states annotated with a transition probability $p(u,v) \in (0,1]$. The primary goal is to identify the most probable path from a designated source node $s$ to a target conversion node $t$.

Figure~\ref{fig:system_workflow} presents the end-to-end workflow. It is organised into five sequential components: (1) data generation and preprocessing, (2) graph construction, (3) baseline path optimization using Dijkstra's algorithm, (4) optimized path search using probability-pruned Dijkstra, and (5) experimental evaluation and comparative analysis.

Two algorithms were implemented and compared throughout this chapter. Algorithm~1 (Baseline Dijkstra on Log-Weights) applies standard Dijkstra's algorithm to the log-probability-transformed graph, guaranteeing globally optimal paths in $O(|E|\log|V|)$ time. Algorithm~2 (Probability-Pruned Dijkstra) extends Algorithm~1 with a pruning threshold $\tau$ that discards partial paths whose cumulative probability falls below $\tau$, reducing expected runtime to $O(|E'|\log|V|)$, $|E'| \leq |E|$, with convergence to Algorithm~1 as $\tau \to 0$.

% ----------------------------------------------------------------
% FIGURE 1: System Workflow
% ----------------------------------------------------------------
\begin{figure}[htbp]
\centering
\begin{tikzpicture}[
    node distance=0.6cm and 0.3cm,
    box/.style={rectangle, rounded corners=4pt, minimum width=2.6cm, minimum height=0.9cm,
                text centered, font=\small\bfseries, draw=black, thick},
    arrow/.style={-{Stealth[length=5pt]}, thick, color=arrowgray},
    label/.style={font=\scriptsize\itshape, text=gray}
]

% Step boxes
\node[box, fill=boxblue]  (s1) {Data Gen.};
\node[box, fill=boxblue, right=of s1]  (s2) {Graph Build};
\node[box, fill=boxorange, right=of s2] (s3) {Baseline Dijkstra};
\node[box, fill=boxgreen,  right=of s3] (s4) {Pruned Dijkstra};
\node[box, fill=boxgray,   right=of s4] (s5) {Evaluation};

% Arrows
\draw[arrow] (s1) -- (s2);
\draw[arrow] (s2) -- (s3);
\draw[arrow] (s3) -- (s4);
\draw[arrow] (s4) -- (s5);

% Sub-labels
\node[label, below=0.15cm of s1] {Synthetic \& Real};
\node[label, below=0.15cm of s2] {$G=(V,E),\; w(u,v)$};
\node[label, below=0.15cm of s3] {$O(|E|\log|V|)$};
\node[label, below=0.15cm of s4] {Threshold $\tau$};
\node[label, below=0.15cm of s5] {Metrics \& Trade-offs};

% Group bracket
\draw[decorate, decoration={brace, amplitude=6pt, mirror}, thick, color=nodeblue]
    ([xshift=-2pt, yshift=-1.1cm]s1.south west) -- ([xshift=2pt, yshift=-1.1cm]s5.south east)
    node[midway, below=8pt, font=\small\bfseries, color=nodeblue]
    {Customer Journey Path Optimization Workflow Process};
\end{tikzpicture}
\caption[System workflow overview]{Five-stage workflow process: data generation, graph construction, baseline Dijkstra, pruned Dijkstra, and experimental evaluation.}
\label{fig:system_workflow}
\end{figure}

\section{Problem Formulation}

\subsection{Graph Model and Inputs}

Customer interactions are modeled as a directed graph $G = (V, E)$, where $V$ is the set of interaction states (nodes), $E \subseteq V \times V$ is the set of directed transitions (edges), and each edge $(u,v) \in E$ carries a transition probability $p(u,v) \in (0,1]$. A designated source node $s \in V$ (e.g., Landing Page) and target node $t \in V$ (e.g., Checkout) define the query endpoints. In the optimized algorithm, a pruning threshold $\tau \in (0,1]$ is used to discard low-probability partial paths.

Figure~\ref{fig:journey_graph} illustrates a representative small customer journey graph to clarify the graph model.

% ----------------------------------------------------------------
% FIGURE 2: Example Customer Journey Graph
% ----------------------------------------------------------------
\begin{figure}[htbp]
\centering
\begin{tikzpicture}[
    state/.style={circle, draw=black, thick, minimum size=1.15cm,
                  font=\small\bfseries, fill=boxblue},
    conv/.style={circle, draw=nodegreen!80!black, very thick, minimum size=1.15cm,
                 font=\small\bfseries, fill=boxgreen},
    src/.style={circle, draw=nodeorange!80!black, very thick, minimum size=1.15cm,
                font=\small\bfseries, fill=boxorange},
    epath/.style={-{Stealth[length=6pt]}, very thick, color=nodeblue},
    earrow/.style={-{Stealth[length=5pt]}, thick, color=black!40, dashed},
    elabel/.style={font=\scriptsize\itshape, fill=white, inner sep=1pt}
]
\node[src]   (LP) at (0,0)     {LP};
\node[state] (SR) at (3.2,0)   {SR};
\node[state] (PP) at (6.4,1.4) {PP};
\node[state] (CT) at (6.4,-1.4){CT};
\node[conv]  (CK) at (9.6,1.4) {CK};

\draw[epath] (LP) -- node[elabel,above]{$p=0.7$} (SR);
\draw[epath] (SR) -- node[elabel,above left]{$p=0.6$} (PP);
\draw[epath] (PP) -- node[elabel,above]{$p=0.8$} (CK);

\draw[earrow] (SR) -- node[elabel,below left]{$p=0.4$} (CT);
\draw[earrow] (CT) to[bend right=18] node[elabel,below right]{$p=0.3$} (CK);
\draw[earrow] (LP) to[bend right=30] node[elabel,below]{$p=0.3$} (CT);
\draw[earrow] (PP) -- node[elabel,right]{$p=0.2$} (CT);

\node[draw=black!30, rounded corners=4pt, fill=white, inner sep=6pt,
      below=1.6cm of SR, xshift=1.8cm] (legbox) {%
  \begin{tabular}{@{}ll@{}}
    \tikz[baseline=-0.5ex]\draw[epath,shorten >=1pt] (0,0)--(0.55,0); &
      Optimal path $(s \!\to\! t)$\\[3pt]
    \tikz[baseline=-0.5ex]\draw[earrow,shorten >=1pt] (0,0)--(0.55,0); &
      Alternative transitions\\[3pt]
    \tikz[baseline=-0.5ex]\node[src,minimum size=0.38cm,inner sep=0pt,font=\tiny]{}; &
      Source $s$\ (LP = Landing Page)\\[3pt]
    \tikz[baseline=-0.5ex]\node[conv,minimum size=0.38cm,inner sep=0pt,font=\tiny]{}; &
      Target $t$\ (CK = Checkout)
  \end{tabular}
};
\end{tikzpicture}
\caption[Example customer journey graph]{Representative customer journey graph. LP = source $s$, CK = target $t$. Solid blue path LP$\to$SR$\to$PP$\to$CK is optimal ($\Pi^*=0.336$, log-cost$\approx1.091$).}
\label{fig:journey_graph}
\end{figure}

\subsection{Probabilistic Path Objective and Outputs}

The goal is to find a path $P^* = (v_0 = s,\, v_1,\, \ldots,\, v_k = t)$ that maximizes the joint path probability~\citep{norris1998markov}:

\[
\max_{P} \prod_{(u,v) \in P} p(u,v)
\]

Directly multiplying many small probabilities causes numerical underflow~\citep{manning1999nlp}. Each transition probability is transformed into an additive non-negative edge weight using the negative logarithm:

\[
w(u,v) = -\log\bigl(p(u,v)\bigr)
\]

Since $p(u,v) \in (0,1]$, it follows that $w(u,v) \geq 0$, satisfying Dijkstra's non-negativity requirement~\citep{cormen2009}:

\[
\max \prod_{(u,v)\in P^*} p(u,v) \;\iff\; \min \sum_{(u,v)\in P^*} -\log\, p(u,v)
\]

For a given source--target query, the system returns the optimal path $P^* = (v_0 = s, v_1, \ldots, v_k = t)$, the corresponding path probability $\Pi^* = \prod_{(u,v)\in P^*} p(u,v) = \exp(-C^*)$, and the log-cost $C^* = \sum_{(u,v)\in P^*} -\log\, p(u,v)$. Each run also records performance metrics---execution time, peak memory, nodes explored, edges relaxed, and maximum priority queue size---as well as the optimality gap $\Delta = |C^*_{\text{baseline}} - C^*_{\text{pruned}}| / C^*_{\text{baseline}} \times 100\%$.

\paragraph{Concrete worked example: RetailRocket event-level funnel.}
The RetailRocket dataset~\citep{retailrocket2015} yields a 3-node event-type graph whose MLE transition probabilities, computed from 2.7\,million events, provide a concrete illustration of the formulation:

\begin{center}
\small
\begin{tabular}{@{}lrl@{}}
\toprule
\textbf{Transition} & \textbf{Probability} & \textbf{Log-weight} \\
\midrule
view $\to$ addtocart    & $p = 0.041$ & $w = -\ln(0.041) = 3.20$ \\
addtocart $\to$ transaction & $p = 0.226$ & $w = -\ln(0.226) = 1.49$ \\
view $\to$ transaction (direct) & $p = 0.004$ & $w = -\ln(0.004) = 5.55$ \\
\bottomrule
\end{tabular}
\end{center}

\noindent Two candidate $s{\to}t$ paths exist from \emph{view} to \emph{transaction}:
\begin{itemize}[nosep]
  \item \textbf{Funnel path:} view $\to$ addtocart $\to$ transaction, $\Pi = 0.041 \times 0.226 = 0.0093$ (log-cost $4.69$).
  \item \textbf{Direct skip:} view $\to$ transaction, $\Pi = 0.004$ (log-cost $5.55$).
\end{itemize}
Dijkstra selects the \emph{funnel path} because it has lower log-cost ($4.69 < 5.55$), equivalently higher probability ($0.93\% > 0.39\%$). This $2.4\times$ probability advantage quantifies the value of the add-to-cart step: users who add items to their cart are substantially more likely to complete a transaction than those who skip directly.

\subsection{Constraints and Assumptions}

The framework assumes graph sizes up to 50,000 nodes (hardware permitting), with experiments conducted on graphs of up to 10,000 nodes. Graphs are assumed to be sparse ($|E| = O(|V|)$) and stored using an adjacency-list representation requiring $O(|V|+|E|)$ space. All transition probabilities are strictly positive; zero-probability transitions are treated as non-existent edges. The customer journey graph may contain directed cycles; correctness is preserved because $w(u,v) \geq 0$ for all edges. Finally, a first-order \textbf{Markov assumption} is adopted: $P(\text{next} \mid \text{current})$ is independent of earlier states. This assumption is standard in clickstream modeling~\citep{bucklin2003clickstream} and is empirically supported by the observation that transition matrices estimated from session data yield stable, reproducible graph structures across independent subsamples. While higher-order dependencies exist (e.g., users arriving via search may navigate differently from those arriving via advertisement), the first-order model captures the dominant transition structure and keeps the state space tractable at $O(|V|)$ rather than $O(|V|^k)$ for order~$k$.

\section{Data Generation and Preprocessing}

\subsection{Synthetic Data Generation}

Two graph generators were implemented in \texttt{data/graph\_generator.py}.

The first is an \textbf{Erd\H{o}s--R\'{e}nyi sparse random generator}~\citep{erdosrenyi1960}, which creates $|V|$ nodes and samples directed edges with expected total $\bar{d} \cdot |V|$; transition probabilities are drawn from either a uniform $\mathcal{U}(0.01, 1.0)$ or a power-law ($\alpha=2$) distribution. This generator produces graphs with uniformly random structure, serving as a controlled baseline for evaluating algorithmic behaviour without structural bias.

The second is a \textbf{layered (stage-based) generator}, which models the funnel-shaped structure typical of real customer journeys. Nodes are distributed across five ordered stages---Awareness, Interest, Consideration, Intent, and Conversion---reflecting the standard e-commerce conversion funnel. Edges are sampled primarily in the forward direction (up to $+2$ stages ahead), ensuring that the dominant flow proceeds from awareness toward conversion. A configurable \texttt{backward\_prob} parameter (default~$0.1$) controls the fraction of backward links, modelling the common user behaviour of revisiting earlier stages (e.g., returning from a product page to browse). Transition probabilities for forward edges are drawn from $\mathcal{U}(0.3, 0.8)$, while backward edges receive lower probabilities from $\mathcal{U}(0.05, 0.3)$, reflecting the empirical observation that forward progression is more likely than backtracking. This layered design is the more application-relevant of the two generators, as it approximates the stage-wise structure observed in real clickstream data (Section~\ref{sec:real_results}).

Both generators employ $O(n \cdot d)$ scalable edge sampling rather than $O(n^2)$ adjacency-matrix construction. A BFS-based connectivity guarantee ensures that if no $s \to t$ path exists after random sampling, bridge edges are added. Outgoing probabilities are normalised per node ($\sum_v p(u,v) = 1$), and deterministic seeding via \texttt{hashlib.md5} ensures cross-platform reproducibility.

The following code snippet shows key aspects of the Erd\H{o}s--R\'{e}nyi generator:

\begin{lstlisting}[language=Python, caption={Core edge sampling from the ER generator (\texttt{data/graph\_generator.py})}, label=lst:er_gen]
def generate_erdos_renyi_graph(n, avg_degree, distribution, seed, ...):
    rng = random.Random(seed)
    graph = {f"node_{i}": {} for i in range(n)}
    # O(n*d) edge sampling -- NOT O(n^2)
    for u in graph:
        num_edges = rng.randint(
            max(0, int(avg_degree) - 2), int(avg_degree) + 2)
        targets = rng.sample(other_nodes, min(num_edges, len(other_nodes)))
        for v in targets:
            if distribution == "uniform":
                raw = rng.uniform(0.01, 1.0)
            else:  # power_law: inverse-CDF Pareto(alpha=2, x_min=0.01)
                raw = min(0.01 * (1 - rng.random()) ** (-1.0), 1.0)
            graph[u][v] = raw
    _normalize_and_log_transform(graph)
    _ensure_connectivity(graph, source, target)
    return graph
\end{lstlisting}

\subsection{Real-World Clickstream Data}

Two real-world clickstream datasets were used for validation (Table~\ref{table:real_datasets}):

\begin{table}[htbp]
\centering
\renewcommand{\arraystretch}{1.3}
\resizebox{\textwidth}{!}{%
\begin{tabular}{@{}lrrrl@{}}
\toprule
\textbf{Dataset} & \textbf{Events} & \textbf{Sessions} & \textbf{Graph Size} & \textbf{Source} \\
\midrule
RetailRocket (event-level) & 2.7M & 1.4M & 3 nodes, 9 edges & Kaggle~\citep{retailrocket2015} \\
RetailRocket (item-level) & 2.7M & 50K & 44,711 nodes, 101,528 edges & Kaggle~\citep{retailrocket2015} \\
RecSys 2015 (YOOCHOOSE) & 33M & 50K & 12,935 nodes, 70,442 edges & Kaggle~\citep{recsys2015} \\
\bottomrule
\end{tabular}%
}
\caption[Real-world dataset summary]{Real-world clickstream datasets used for validation. RetailRocket supports two granularities: event-type (3-node funnel) and item-level (large graph). RecSys 2015 uses item-to-item click transitions.}
\label{table:real_datasets}
\end{table}

Transition probabilities are estimated using maximum likelihood:
\[
\hat{p}(u,v) = \frac{\text{count}(u \to v)}{\displaystyle\sum_{w:\,(u,w)\in E} \text{count}(u \to w)}
\]

\noindent A consequence of MLE on sparse data is that nodes with very few observed transitions (e.g., items viewed only once or twice) receive high-variance probability estimates. In item-level graphs, where tens of thousands of items may each have only a handful of observations, many edges carry very low estimated probabilities. This contributes to the low path-found rates observed in item-level real-data experiments (Section~\ref{sec:real_results}): when most edge probabilities are small, even the baseline optimal path has a very low cumulative probability, making it more susceptible to pruning. Smoothing techniques such as Laplace (add-one) correction or Bayesian priors could mitigate this effect but were not applied in order to isolate the algorithmic contribution of pruning.

\subsection{Preprocessing and Input Formats}

The preprocessing module (\texttt{src/preprocessing.py}) accepts two CSV formats:

\begin{enumerate}
    \item \textbf{Direct transitions} --- columns \texttt{source} and \texttt{target}, one row per observed transition.
    \item \textbf{Session event streams} --- columns \texttt{session\_id} and \texttt{state} (plus \texttt{step} or \texttt{timestamp} for ordering). Consecutive events within each session are converted to pairwise transitions.
\end{enumerate}

Both formats produce the same transition-count table from which MLE probabilities are computed.

\section{Graph Construction}

The directed weighted graph is constructed from the computed transition probabilities using an \textbf{adjacency list} representation. The function receives a nested dictionary mapping each source node to its targets and their probabilities, then converts each probability into a log-weight $w(u,v) = -\log(\hat{p}(u,v))$. This provides $O(|V|+|E|)$ space complexity and efficient $O(\deg(u))$ neighbour iteration~\citep{cormen2009}.

\begin{lstlisting}[language=Python, caption={Graph construction (\texttt{src/graph\_builder.py})}, label=lst:graph_build]
def build_weighted_graph(transition_probs):
    graph = defaultdict(list)
    for source, targets in transition_probs.items():
        for target, prob in targets.items():
            if prob <= 0:
                continue
            graph[source].append((target, -math.log(prob)))
            if target not in graph:
                graph[target] = []  # ensure sink nodes exist
    return dict(graph)
\end{lstlisting}

\section{Algorithmic Approaches}

\subsection{Baseline Algorithm: Dijkstra's Shortest Path}

The baseline applies Dijkstra's algorithm~\citep{dijkstra1959} to the log-transformed graph. A \textbf{binary min-heap} (\texttt{heapq}) serves as the priority queue with \textbf{lazy insertion}---when a shorter path to a node is found, a new entry is pushed and stale entries are skipped when popped.

\begin{algorithm}[H]
\caption{Dijkstra on Transformed Probabilities~\citep{dijkstra1959,cormen2009}}\label{alg:baseline}
\SetAlgoLined
\KwIn{Graph $G=(V,E)$ with weights $w(u,v)=-\log p(u,v)$, source $s$, target $t$}
\KwOut{Shortest-path distance $\mathit{dist}[t]$, probability $\exp(-\mathit{dist}[t])$, path $P^*$}
Initialise $\mathit{dist}[v] \leftarrow \infty$ for all $v \in V$; $\mathit{parent}[v] \leftarrow \mathrm{NIL}$\;
$\mathit{dist}[s] \leftarrow 0$; insert $(0, s)$ into min-heap $Q$\;
\While{$Q$ is not empty}{
  $(d, u) \leftarrow \mathrm{Extract\text{-}Min}(Q)$\;
  \If{$u = t$}{\textbf{break}\tcp*{optimal path reached}}
  \If{$d > \mathit{dist}[u]$}{\textbf{continue}\tcp*{stale entry (lazy deletion)}}
  \ForEach{edge $(u, v)$ with weight $w = -\log p(u,v)$}{
    $\mathit{edges\_relaxed} \leftarrow \mathit{edges\_relaxed} + 1$\;
    $\mathit{new\_dist} \leftarrow \mathit{dist}[u] + w$\;
    \If{$\mathit{new\_dist} < \mathit{dist}[v]$}{
      $\mathit{dist}[v] \leftarrow \mathit{new\_dist}$; $\mathit{parent}[v] \leftarrow u$\;
      Insert $(\mathit{new\_dist}, v)$ into $Q$\;
    }
  }
}
\Return $\mathit{dist}[t]$, $\exp(-\mathit{dist}[t])$, $\mathrm{ReconstructPath}(\mathit{parent}, s, t)$\;
\end{algorithm}

\noindent\textbf{Time complexity:} $O(|E| \log |V|)$ with a binary heap~\citep{cormen2009}.\\
\textbf{Space complexity:} $O(|V| + |E|)$.\\
\textbf{Optimality:} Guaranteed, since all $w(u,v) \geq 0$~\citep{dijkstra1959}.

\subsubsection{Worked Example}

Table~\ref{table:dijkstra_trace} traces Algorithm~1 on the five-node graph (Figure~\ref{fig:journey_graph}).

\begin{table}[htbp]
\centering
\begin{tabular}{@{}ccccccc@{}}
\toprule
\textbf{Step \#} & \textbf{Extract} & \textbf{LP} & \textbf{SR} & \textbf{PP} & \textbf{CT} & \textbf{CK} \\
\midrule
Init   & ---          & $0$    & $\infty$ & $\infty$ & $\infty$ & $\infty$ \\
1      & LP           & $0$    & $0.36$   & $\infty$ & $1.20$   & $\infty$ \\
2      & SR           & $0$    & $0.36$   & $0.87$   & $1.20$   & $\infty$ \\
3      & PP           & $0$    & $0.36$   & $0.87$   & $1.20$   & $1.09$ \\
4      & \textbf{CK}\checkmark & $0$ & $0.36$ & $0.87$ & $1.20$ & \textbf{1.09} \\
\bottomrule
\end{tabular}
\caption[Dijkstra dist trace]{Dijkstra \texttt{dist[]} trace on the five-node graph, $s = \text{LP}$, $t = \text{CK}$. At step~4, CK is finalised with optimal log-cost $1.09$, probability $\exp(-1.09) = 0.336$.}
\label{table:dijkstra_trace}
\end{table}

\subsection{Optimized Algorithm: Probability-Pruned Dijkstra}

The pruned algorithm extends the baseline by introducing a \textbf{probability pruning threshold} $\tau \in (0, 1]$. Any partial path whose cumulative probability falls below $\tau$ is pruned. In log-space:

\[
\text{new\_cost} > T = -\log(\tau) \implies \text{skip edge}
\]

\begin{algorithm}[H]
\caption{Probability-Pruned Dijkstra}\label{alg:pruned}
\SetAlgoLined
\KwIn{Graph $G=(V,E)$, source $s$, target $t$, pruning threshold $\tau \in (0,1]$}
\KwOut{Shortest-path distance $\mathit{dist}[t]$, probability $\exp(-\mathit{dist}[t])$, path $P^*$}
Initialise $\mathit{dist}[v] \leftarrow \infty$ for all $v \in V$; $\mathit{parent}[v] \leftarrow \mathrm{NIL}$\;
$\mathit{dist}[s] \leftarrow 0$; insert $(0, s)$ into min-heap $Q$\;
$T \leftarrow -\log(\tau)$\tcp*{pruning threshold in log-space}
\While{$Q$ is not empty}{
  $(d, u) \leftarrow \mathrm{Extract\text{-}Min}(Q)$\;
  \If{$u = t$}{\textbf{break}}
  \If{$d > \mathit{dist}[u]$}{\textbf{continue}\tcp*{stale entry}}
  \ForEach{edge $(u, v)$ with weight $w = -\log p(u,v)$}{
    $\mathit{new\_dist} \leftarrow \mathit{dist}[u] + w$\;
    \If{$\mathit{new\_dist} > T$}{\textbf{continue}\tcp*{PRUNE: below $\tau$}}
    $\mathit{edges\_relaxed} \leftarrow \mathit{edges\_relaxed} + 1$\;
    \If{$\mathit{new\_dist} < \mathit{dist}[v]$}{
      $\mathit{dist}[v] \leftarrow \mathit{new\_dist}$; $\mathit{parent}[v] \leftarrow u$\;
      Insert $(\mathit{new\_dist}, v)$ into $Q$\;
    }
  }
}
\Return $\mathit{dist}[t]$, $\exp(-\mathit{dist}[t])$, $\mathrm{ReconstructPath}(\mathit{parent}, s, t)$\;
\end{algorithm}

\noindent\textbf{Worst-case time complexity:} $O(|E| \log |V|)$ (when $\tau \to 0$).\\
\textbf{Empirically observed time complexity:} $O(|E'| \log |V|)$, $|E'| \leq |E|$ (see Chapter~\ref{chap:complexity} for validation).\\
\textbf{Space complexity:} $O(|V| + |E|)$.

\subsubsection{The \texttt{edges\_relaxed} Metric}

The \texttt{edges\_relaxed} counter has a precise, intentional definition that differs between the two algorithms:

\begin{table}[htbp]
\centering
\renewcommand{\arraystretch}{1.3}
\begin{tabular}{@{}p{3cm}p{5cm}p{5.5cm}@{}}
\toprule
\textbf{Algorithm} & \textbf{Counts} & \textbf{Placement in Code} \\
\midrule
Baseline & Every outgoing edge examined from a settled node & After computing \texttt{new\_cost}, before the improvement check \\
Pruned & Every edge that \textbf{survived} the $\tau$ threshold & After the \texttt{if new\_cost > T: continue} check \\
\bottomrule
\end{tabular}
\caption[edges\_relaxed semantics]{The \texttt{edges\_relaxed} metric definition. Because the counter is placed at different points in the two algorithms, the baseline count includes all edge examinations whereas the pruned count includes only edges that survived the threshold check. This is an intentional design choice: the ratio $\text{baseline}/\text{pruned}$ directly quantifies the computational work eliminated by pruning, though readers should note that the two counts are not strictly comparable in absolute terms.}
\label{table:edges_relaxed}
\end{table}

This design cleanly isolates the pruning effect from stale-entry skips, making the speedup ratio directly attributable to the $\tau$ mechanism.

\subsubsection{Actual Python Implementation}

The following is the actual pruned Dijkstra implementation from the codebase:

\begin{lstlisting}[language=Python, caption={Pruned Dijkstra implementation (\texttt{src/dijkstra.py})}, label=lst:pruned_impl]
def dijkstra_pruned(graph, source, target, tau):
    T = -math.log(tau) if tau > 0 else float("inf")
    dist = {source: 0.0}
    parent = {source: None}
    pq = [(0.0, source)]
    metrics = DijkstraMetrics()

    while pq:
        d, u = heapq.heappop(pq)
        if u == target:
            break
        if d > dist.get(u, float("inf")):
            continue  # stale entry
        metrics.nodes_explored += 1
        metrics.max_pq_size = max(metrics.max_pq_size, len(pq)+1)
        for v, w in graph.get(u, []):
            new_cost = d + w
            if new_cost > T:
                continue  # PRUNE
            metrics.edges_relaxed += 1
            if new_cost < dist.get(v, float("inf")):
                dist[v] = new_cost
                parent[v] = u
                heapq.heappush(pq, (new_cost, v))
    # ... path reconstruction and probability computation
\end{lstlisting}

\subsection{Algorithm Comparison}

Figure~\ref{fig:algo_flowchart} illustrates the decision-flow difference between both algorithms. Table~\ref{table:algo_compare} summarises the trade-offs.

% ----------------------------------------------------------------
% FIGURE 3: Algorithm Comparison Flowchart
% ----------------------------------------------------------------
\begin{figure}[htbp]
\centering
\resizebox{\textwidth}{!}{%
\begin{tikzpicture}[
    node distance=0.5cm and 1.6cm,
    startstop/.style={rectangle, rounded corners=8pt, minimum width=2.5cm, minimum height=0.7cm,
                      draw=black, thick, font=\small\bfseries, text centered},
    process/.style={rectangle, minimum width=2.5cm, minimum height=0.7cm,
                    draw=black, thick, font=\small, text centered},
    decision/.style={diamond, aspect=1.8, minimum width=2.8cm, minimum height=0.6cm,
                     draw=black, thick, font=\scriptsize\bfseries, text centered},
    prune/.style={rectangle, minimum width=2.5cm, minimum height=0.7cm,
                  draw=red!70!black, thick, fill=boxred, font=\small, text centered},
    arr/.style={-{Stealth[length=4pt]}, thick},
    lbl/.style={font=\scriptsize, fill=white, inner sep=1pt}
]

%--- LEFT COLUMN: Baseline Dijkstra ---
\node[startstop, fill=boxorange] (b_start) {Extract-Min$(Q)$};
\node[decision, below=0.5cm of b_start]  (b_tgt)   {$u = t$?};
\node[decision, below=0.7cm of b_tgt]    (b_stale)  {Stale?};
\node[process,  below=0.7cm of b_stale, fill=boxblue] (b_relax) {Relax neighbours};
\node[process,  below=0.5cm of b_relax, fill=boxblue] (b_push)  {Push to $Q$ if better};

\draw[arr] (b_start) -- (b_tgt);
\draw[arr] (b_tgt)   -- node[lbl,right]{No}  (b_stale);
\draw[arr] (b_stale) -- node[lbl,right]{No}  (b_relax);
\draw[arr] (b_relax) --                      (b_push);

\draw[arr] (b_tgt.east)
    -- ++(1.1,0)
    |- node[lbl, pos=0.25, right]{Yes $\Rightarrow$ \textbf{Stop}}
    (b_push.east);

\draw[arr] (b_stale.east)
    -- ++(0.8,0)
    |- node[lbl, pos=0.25, right]{Yes $\Rightarrow$ Skip}
    (b_relax.east);

\node[font=\small\bfseries, above=0.25cm of b_start, color=nodeorange] {Baseline Dijkstra};

%--- RIGHT COLUMN: Pruned Dijkstra ---
\node[startstop, fill=boxorange, right=3.8cm of b_start] (p_start) {Extract-Min$(Q)$};
\node[decision,  below=0.5cm of p_start]  (p_tgt)   {$u = t$?};
\node[decision,  below=0.7cm of p_tgt]    (p_stale)  {Stale?};
\node[decision,  below=0.7cm of p_stale]  (p_prune)  {$d_{\text{new}} > T$?};
\node[prune,     below=0.5cm of p_prune]  (p_skip)   {\textbf{PRUNE} (skip)};
\node[process,   right=1.3cm of p_prune, fill=boxblue] (p_relax)  {Relax edge};
\node[process,   below=0.5cm of p_skip,  fill=boxblue] (p_push)   {Push to $Q$ if better};

\draw[arr] (p_start) -- (p_tgt);
\draw[arr] (p_tgt)   -- node[lbl,right]{No}  (p_stale);
\draw[arr] (p_stale) -- node[lbl,right]{No}  (p_prune);
\draw[arr] (p_prune) -- node[lbl,left] {Yes} (p_skip);
\draw[arr] (p_skip)  --                      (p_push);

\draw[arr] (p_prune.east) -- node[lbl,above]{No} (p_relax);
\draw[arr] (p_relax.south) |- (p_push.east);

\draw[arr] (p_tgt.east)
    -- ++(1.4,0)
    |- node[lbl, pos=0.25, right]{Yes $\Rightarrow$ \textbf{Stop}}
    (p_push.east);

\draw[arr] (p_stale.east)
    -- ++(1.1,0)
    |- node[lbl, pos=0.25, right]{Yes $\Rightarrow$ Skip}
    (p_prune.east);

\node[font=\small\bfseries, above=0.25cm of p_start, color=nodegreen] {Pruned Dijkstra};
\end{tikzpicture}%
}% end resizebox
\caption[Algorithm decision-flow comparison]{Decision-flow comparison. The pruned variant adds a threshold check $d_{\text{new}}>T$ (red box); edges exceeding $T=-\log(\tau)$ are discarded without relaxation.}
\label{fig:algo_flowchart}
\end{figure}

\begin{table}[htbp]
\centering
\renewcommand{\arraystretch}{1.3}
\begin{tabular}{@{}p{2.8cm}p{5.5cm}p{5.5cm}@{}}
\toprule
\textbf{Criterion} & \textbf{Baseline Dijkstra} & \textbf{Probability-Pruned Dijkstra} \\
\midrule
\textbf{Optimality} & Guaranteed globally optimal & Optimal when path cost $\leq T$; otherwise no path returned \\[4pt]
\textbf{Time} & $O(|E|\log|V|)$ & $O(|E'|\log|V|)$, $|E'| \leq |E|$ \\[4pt]
\textbf{Space} & $O(|V|+|E|)$ & $O(|V|+|E|)$ (unchanged) \\[4pt]
\textbf{Trade-off} & None & Reachability vs.\ speed \\[4pt]
\textbf{Convergence} & Always optimal & Identical to baseline as $\tau \to 0$ \\
\bottomrule
\end{tabular}
\caption[Algorithm comparison]{Comparison of Baseline and Pruned Dijkstra. Experimental results (Chapter~\ref{chap:results}) confirm 0\% optimality gap on all found paths.}
\label{table:algo_compare}
\end{table}

\subsection{Critical-$\tau$ Finder}
\label{sec:critical_tau}

A practical question is: \emph{what is the largest threshold $\tau^*$ that still preserves the optimal path?} The module \texttt{src/critical\_tau.py} answers this with an adaptive sweep centred on the baseline path probability.

Given the baseline optimal probability $\Pi^*$, fixed $\tau$ values would miss the interesting region when $\Pi^*$ is very small (e.g.\ $10^{-4}$). Instead, the sweep generates $\tau$ candidates as fractions of $\Pi^*$---specifically $\{0.10, 0.30, 0.50, 0.70, 0.90, 0.95, 0.99, 1.00, 1.10\} \times \Pi^*$. The $110\%$ probe intentionally exceeds the optimal probability to identify the \emph{cliff point} where the pruned algorithm first fails to find a path. For each candidate, the pruned algorithm is timed, and speedup (both node-count and wall-clock) and optimality gap are recorded. The critical threshold $\tau^*$ is the largest tested $\tau$ for which the gap remains below a user-defined tolerance (default 1\%).

\subsection{$k$-Shortest Simple Paths}

In addition to single-path queries, the CLI (\texttt{main.py}) provides a $k$-shortest simple paths mode activated by the \texttt{--k} flag. This uses a best-first search with visited-set tracking to enumerate the top-$k$ distinct simple paths from $s$ to $t$ in non-decreasing cost order. When $k=1$, the result is identical to Dijkstra's shortest path. This feature is validated by four dedicated unit tests covering $k=1$ equivalence, ascending cost ordering, simple-path guarantee, and disconnected-graph behaviour.

\section{Data Structures}

\subsection{Adjacency List}

The directed graph is stored as a nested Python dictionary mapping each node $u$ to its outgoing neighbours and associated log-transformed weights. This ensures $O(|V|+|E|)$ space and $O(\deg(u))$ neighbour iteration.

\subsection{Binary Min-Heap (Priority Queue)}

Python's \texttt{heapq} module provides $O(\log|V|)$ extract-min and insert. Since \texttt{heapq} does not support decrease-key, \textbf{lazy insertion} is used: when a shorter path to a node is found, a new entry is pushed; stale entries are skipped when popped (the \texttt{if d > dist[u]: continue} check).

\subsection{Hash Maps (Dictionaries)}

\texttt{dist[v]} and \texttt{parent[v]} are Python dictionaries with $O(1)$ average-case access.

\section{Experimental Design}
\label{sec:expdesign}

\subsection{Simulation Environment}

Table~\ref{table:simenv} summarises the environment used for all experiments.

\begin{table}[htbp]
\centering
\renewcommand{\arraystretch}{1.3}
\begin{tabular}{@{}ll@{}}
\toprule
\textbf{Component} & \textbf{Specification} \\
\midrule
Programming Language & Python 3.13.3 (standard library only for core algorithms) \\
Graph Representation & Adjacency list (\texttt{dict} of \texttt{dict}) \\
Priority Queue & \texttt{heapq} (binary min-heap with lazy insertion) \\
Runtime Measurement & \texttt{time.perf\_counter()} (separate pass from memory) \\
Memory Measurement & \texttt{tracemalloc} peak allocation (separate pass from timing) \\
Random Seeds & Deterministic \texttt{hashlib.md5} per configuration \\
Repetitions & 10 runs per configuration \\
Graph Generators & Erd\H{o}s--R\'{e}nyi + Layered (in-house, \texttt{data/graph\_generator.py}) \\
\bottomrule
\end{tabular}
\caption[Simulation environment]{Simulation environment. Timing and memory are measured in separate passes so \texttt{tracemalloc} overhead does not corrupt the stopwatch.}
\label{table:simenv}
\end{table}

\subsection{Evaluation Metrics}

Each experiment records execution time (wall-clock milliseconds via \texttt{time.perf\_counter()}), peak memory (bytes via \texttt{tracemalloc.get\_traced\_memory()}), nodes explored (vertices extracted from the priority queue), edges relaxed (edge operations; see Table~\ref{table:edges_relaxed}), maximum priority queue size, and path probability $\Pi^* = \exp(-C^*)$. The optimality gap is defined as:
    \[
    \Delta = \frac{|C^*_{\text{baseline}} - C^*_{\text{pruned}}|}{C^*_{\text{baseline}}} \times 100\%
    \]
    This cost-based comparison (rather than probability-based) eliminates floating-point rounding noise.

\subsection{Experimental Parameters}

\begin{table}[htbp]
\centering
\renewcommand{\arraystretch}{1.3}
\begin{tabular}{@{}llp{7cm}@{}}
\toprule
\textbf{Parameter} & \textbf{Values} & \textbf{Purpose} \\
\midrule
Graph type & ER, Layered & Test generic vs.\ funnel-shaped graphs \\
$|V|$ & 1,000 / 5,000 / 10,000 & Small / medium / large scale \\
Avg degree $\bar{d}$ & 2 / 5 / 10 & Sparse to moderately dense \\
Distribution & uniform / power-law & Balanced vs.\ heterogeneous transitions \\
$\tau$ & 0, 0.001, 0.01, 0.05, 0.1, 0.5 & Conservative to aggressive pruning \\
Runs & 10 per config & Statistical reliability \\
\bottomrule
\end{tabular}
\caption[Experimental parameter matrix]{Full parameter matrix: $2 \times 3 \times 3 \times 2 \times 6 \times 10 = 2{,}160$ total rows.}
\label{table:param_matrix}
\end{table}

The $\tau$ values were chosen to span three orders of magnitude, from conservative ($\tau = 0.001$, retaining paths with cumulative probability $\geq 0.1\%$) to aggressive ($\tau = 0.5$, retaining only paths with $\geq 50\%$ probability). The intermediate values $\{0.01, 0.05, 0.1\}$ sample the transition region where speedup gains accelerate but path-found rates begin to drop. $\tau = 0$ serves as the control condition (no pruning, identical to baseline). For practitioners seeking to choose $\tau$ on a specific graph, the \texttt{critical\_tau.py} module (Section~\ref{sec:critical_tau}) provides an adaptive sweep that probes fractions of the baseline optimal probability $\Pi^*$ to identify the largest $\tau$ that preserves the optimal path.

\subsection{Experimental Controls}

To ensure fair comparison, both algorithms run on identical graph instances using the same \texttt{hashlib.md5} seed. Execution is single-threaded to eliminate scheduling noise, and each configuration is repeated 10 times with median values reported (median is preferred over mean for skewed runtime distributions). Timing and memory are measured in separate passes so that \texttt{tracemalloc} overhead does not affect wall-clock measurements.
