\chapter{CORRECTNESS PROOFS AND VALIDATION}\label{chap:correctness}

This chapter provides formal correctness proofs for both algorithms and presents the empirical validation results that confirm theoretical guarantees. Theorems are numbered sequentially within this chapter (Theorems~\ref{thm:log_transform}--\ref{thm:all_or_nothing}); all cross-references in the report use labelled theorem names to avoid ambiguity.

\section{Correctness of the Log-Probability Transformation}

\begin{theorem}[Order-Preserving Log Transform]\label{thm:log_transform}
The path $P^*$ that minimizes $\sum_{(u,v)\in P} -\log\,p(u,v)$ is identical to the path that maximizes $\prod_{(u,v)\in P} p(u,v)$.
\end{theorem}

\begin{proof}
The negative logarithm $f(x)=-\log(x)$ is a strictly monotone decreasing function on $(0,1]$~\citep{cormen2009}. For any two paths $P_1,P_2$ from $s$ to $t$:
\begin{align}
\prod_{(u,v)\in P_1} p(u,v) > \prod_{(u,v)\in P_2} p(u,v) &\iff \log\!\Bigl(\prod_{(u,v)\in P_1} p(u,v)\Bigr) > \log\!\Bigl(\prod_{(u,v)\in P_2} p(u,v)\Bigr) \nonumber\\
&\iff \sum_{(u,v)\in P_1} \log\,p(u,v) > \sum_{(u,v)\in P_2} \log\,p(u,v) \nonumber\\
&\iff \sum_{(u,v)\in P_1} -\log\,p(u,v) < \sum_{(u,v)\in P_2} -\log\,p(u,v) \nonumber
\end{align}

Since $p(u,v)\in(0,1]$, we have $w(u,v)=-\log\,p(u,v)\geq 0$, satisfying Dijkstra's non-negativity requirement. The transformation is order-preserving: maximizing path probability is equivalent to minimizing log-cost~\citep{manning1999nlp,viterbi1967error}.
\end{proof}

\section{Correctness of Baseline Dijkstra}\label{sec:correctness_baseline}

\begin{theorem}[Baseline Dijkstra Optimality]\label{thm:baseline_optimal}
When Algorithm~1 terminates, $\mathit{dist}[t]$ equals the true shortest-path distance $\delta(s,t)$ in the log-transformed graph.
\end{theorem}

\begin{proof}
The proof proceeds by induction on the number of nodes extracted from the priority queue~\citep{cormen2009,dijkstra1959}.

\textbf{Invariant.} At all times, for every extracted node $u$: $\mathit{dist}[u] = \delta(s,u)$.

\textbf{Base case.} The source $s$ is extracted first with $\mathit{dist}[s]=0=\delta(s,s)$.

\textbf{Inductive step.} Assume the invariant holds for all $k$ previously extracted nodes. Let $u$ be the $(k+1)$-th node extracted with tentative distance $\mathit{dist}[u]$. Suppose for contradiction that there exists a shorter path $s \leadsto x \leadsto u$ where $x$ is not yet extracted. Since all edge weights are non-negative:
\begin{enumerate}
    \item[(i)] $\delta(s,x) \geq \mathit{dist}[u]$ (otherwise $x$ would have been extracted before $u$).
    \item[(ii)] $\delta(s,x) + w(x,u) \geq \delta(s,x) \geq \mathit{dist}[u]$.
\end{enumerate}
This contradicts the assumption that a shorter path exists through $x$. Therefore $\mathit{dist}[u] = \delta(s,u)$.

\textbf{Termination.} The algorithm terminates when $t$ is extracted (early-exit) or the queue becomes empty. The invariant guarantees $\mathit{dist}[t] = \delta(s,t)$ at extraction.

\textbf{Path reconstruction.} Each \texttt{parent[v]} pointer is set only when a strictly shorter path to $v$ is found, forming a shortest-path tree. Backtracking from $t$ to $s$ via parent pointers yields $P^*$.
\end{proof}

\section{Correctness of Probability-Pruned Dijkstra}

\subsection{Convergence at $\tau = 0$}

\begin{theorem}[Convergence]\label{thm:convergence}
When $\tau = 0$, Algorithm~2 is identical to Algorithm~1.
\end{theorem}

\begin{proof}
When $\tau = 0$, the implementation sets $T = \mathtt{math.inf}$ (i.e.\ $T = +\infty$), since $\lim_{\tau\to 0^+}(-\log\tau) = +\infty$. The pruning condition $\mathit{new\_cost} > T$ is never satisfied since all finite costs are $< +\infty$. Therefore no edge is pruned, and every code path of Algorithm~2 behaves identically to Algorithm~1. Correctness follows directly from Theorem~\ref{thm:baseline_optimal}.
\end{proof}

\subsection{Conditional Optimality for $\tau > 0$}

\begin{theorem}[Conditional Optimality]\label{thm:conditional}
If the optimal path $P^*$ has probability $\Pi^* = \prod_{(u,v)\in P^*}p(u,v) \geq \tau$, then Algorithm~2 returns $P^*$ with $\Delta = 0\%$.
\end{theorem}

\begin{proof}
If $\Pi^* \geq \tau$, then $C^* = -\log(\Pi^*) \leq -\log(\tau) = T$. Since Dijkstra processes paths in non-decreasing cost order~\citep{cormen2009}, every prefix of $P^*$ has cost $\leq C^* \leq T$. No edge on the optimal path triggers the pruning condition. The algorithm therefore explores the exact same set of optimal-path nodes as the baseline, and by Theorem~\ref{thm:baseline_optimal}, returns $P^*$ with optimality gap $\Delta = 0\%$.
\end{proof}

\subsection{All-or-Nothing Property}

\begin{theorem}[All-or-Nothing]\label{thm:all_or_nothing}
Algorithm~2 either returns the globally optimal path (identical to the baseline) or reports no path found. It never returns a sub-optimal path.
\end{theorem}

\begin{proof}
Suppose Algorithm~2 returns a path $P'$ from $s$ to $t$. Since the pruning check ($\mathit{new\_cost} > T$) only skips edges---it does not alter edge weights or the priority queue ordering---the settled-node invariant from Theorem~\ref{thm:baseline_optimal} holds for all nodes explored by the pruned algorithm. Therefore, among all paths that survive the threshold, $P'$ minimizes log-cost. If $P^*$ also survives (Theorem~\ref{thm:conditional}), then $P' = P^*$. If $P^*$ is pruned, then $t$ is unreachable in the pruned subgraph and no path is returned.

In either case, the algorithm never returns a sub-optimal path: the gap $\Delta$ is exactly 0\% whenever a path is found.
\end{proof}

\section{Empirical Validation Results}

All five correctness properties were verified experimentally.

Across 2,160 synthetic experiment rows and 840 real-data rows (300 fixed-$\tau$ + 540 adaptive-$\tau$), the optimality gap was measured as $\Delta = |C^*_{\text{baseline}} - C^*_{\text{pruned}}| / C^*_{\text{baseline}} \times 100\%$. The result was $\Delta = 0.0000\%$ for every configuration in which both algorithms found a path (178 of 1,800 pruned experiment rows with $\tau > 0$; 528 of 840 real-data rows), confirming Theorem~\ref{thm:all_or_nothing} empirically.

Convergence testing showed that at $\tau = 0.001$ the pruned algorithm found paths in 30.6\% of configurations, decreasing to 5.6\% at $\tau = 0.01$ and 4.2\% at $\tau = 0.5$. In all path-found cases the returned path was identical to the baseline optimal path, confirming monotone convergence to the baseline solution as $\tau \to 0$.

Probability consistency was verified by comparing $\exp(-C^*)$ against $\prod_{(u,v)\in P^*} p(u,v)$ for all returned paths; relative error remained below $10^{-12}$ in every case.

The test suite includes 41 automated tests covering disconnected source--target pairs (pruned returns \texttt{None}), single-node graphs, complete graphs and star topologies, real-world graphs with high sparsity, and $\tau=0$ equivalence (Theorem~\ref{thm:convergence}). All 41 tests pass.

Finally, the five-node example graph (Figure~\ref{fig:journey_graph}) was traced manually:
\[
C^*(\text{LP}\to\text{SR}\to\text{PP}\to\text{CK}) = -\log(0.7)-\log(0.6)-\log(0.8) \approx 1.091
\]
Both algorithms returned this path and cost, confirming correctness on the reference instance.

\begin{table}[htbp]
\centering
\renewcommand{\arraystretch}{1.3}
\begin{tabular}{@{}p{3.8cm}p{4cm}p{5.5cm}@{}}
\toprule
\textbf{Property} & \textbf{Theoretical} & \textbf{Empirical Confirmation} \\
\midrule
Log transform (Thm~\ref{thm:log_transform}) & Order-preserving & All paths consistent ($<10^{-12}$ error) \\
Baseline optimal (Thm~\ref{thm:baseline_optimal}) & $\mathit{dist}[t]=\delta(s,t)$ & 3,000 rows verified \\
Convergence (Thm~\ref{thm:convergence}) & Identical at $\tau=0$ & Byte-identical output \\
Conditional optimal (Thm~\ref{thm:conditional}) & $\Delta=0$ when $\Pi^*\geq\tau$ & $\Delta=0.0000\%$ on all found paths \\
All-or-nothing (Thm~\ref{thm:all_or_nothing}) & Never sub-optimal & 0 sub-optimal results in 3,000 rows \\
\bottomrule
\end{tabular}
\caption[Correctness validation summary]{Correctness validation summary.}
\label{table:correctness_summary}
\end{table}

\section{Separation of Theoretical and Empirical Claims}

\textit{This section distinguishes between formally proven results, empirically validated findings, and acknowledged limitations.}

\subsection{Fully Proven Claims}

The following statements are mathematical theorems with formal proofs, independent of experimental data:

\begin{enumerate}
\item[(T1)] The log-probability transformation $w(u,v) = -\log p(u,v)$ preserves path ordering (Theorem~\ref{thm:log_transform}).
\item[(T2)] Baseline Dijkstra returns the globally optimal path in $O(|E|\log|V|)$ time (Theorem~\ref{thm:baseline_optimal}).
\item[(T3)] When $\tau = 0$, Algorithm~2 is identical to Algorithm~1 (Theorem~\ref{thm:convergence}).
\item[(T4)] When the optimal path's probability satisfies $\Pi^* \geq \tau$, Algorithm~2 returns it with zero gap (Theorem~\ref{thm:conditional}).
\item[(T5)] Algorithm~2 never returns a sub-optimal path; it either returns the global optimum or reports no path (Theorem~\ref{thm:all_or_nothing}).
\end{enumerate}

\subsection{Empirically Validated Findings}

The following results are supported by experimental data but **not formally guaranteed** for arbitrary graphs:

\begin{enumerate}
\item[(E1)] Wall-clock speedups range from $3.4\times$ ($\tau=0.001$) to $735.2\times$ ($\tau=0.5$) on synthetic graphs (Table~\ref{table:speedup_tau}, Chapter~\ref{chap:results}).
\item[(E2)] Speedup scales monotonically with graph size $|V|$ at fixed $\tau$ (Table~\ref{table:speedup_size}).
\item[(E3)] Erd\H{o}s--R\'{e}nyi graphs exhibit higher speedups than layered graphs at identical $\tau$ (Table~\ref{table:speedup_graph_type}).
\item[(E4)] Critical $\tau_c$ exists in a narrow empirical range where speedup peaks without loss of reachability (Chapter~\ref{chap:results}, Section~5.5).
\item[(E5)] On real-world datasets (RetailRocket, RecSys~2015), speedups reach $18{,}882\times$ and $2{,}309\times$, respectively, under fixed $\tau$ (Chapter~\ref{chap:results}, Section~5.6).
\item[(E6)] Optimality gap remains exactly $0.0000\%$ across 706 path-found cases in combined synthetic and real experiments (all tested configurations).
\end{enumerate}

\subsection{Acknowledged Limitations}

The following limitations are explicitly recognized:

\begin{enumerate}
\item[(L1)] No formal approximation guarantee exists for $\tau > 0$. While we prove convergence (\textit{T3, T4}), the pruned algorithm's behavior for a specific $\tau$ on an arbitrary graph is not formally characterized. Speedup and reachability trade-offs are purely empirical.
\item[(L2)] Critical $\tau_c$ is defined empirically; we provide no theoretical method to predict it a priori for a given graph without experimental sweep.
\item[(L3)] Results assume edge weights are non-negative (the log-probability transformation guarantees this, by construction). The framework does not extend to negative weights or negative-cycle detection.
\item[(L4)] Real-world validation is limited to two e-commerce clickstream datasets. Generalization to other domains (social networks, transportation, etc.) is not formally justified and would require further empirical testing.
\end{enumerate}
