\chapter{CONCLUSION AND FUTURE WORK}\label{chap:conclusion}

\section{Summary of Contributions}

This project investigated the effectiveness of probability-based pruning for shortest-path search in customer journey graphs. Two algorithms were implemented, tested, and compared: a baseline Dijkstra algorithm on log-probability-transformed graphs ($O(|E|\log|V|)$) and a probability-pruned Dijkstra variant with a tunable threshold $\tau$ ($O(|E'|\log|V|)$, $|E'|\leq|E|$).

The key contributions are summarized below.

The pruning mechanism yields substantial speedups with zero quality loss. On synthetic benchmarks (2,160 rows), wall-clock speedups range from $3.4\times$ at $\tau=0.001$ to $735.2\times$ at $\tau=0.5$, with edge-relaxation speedups reaching $1{,}928\times$ (fair edge-examination speedups up to $710\times$). On real-world item-level graphs under fixed $\tau$, a maximum speedup of $18{,}882\times$ was recorded---reducing query time from ${\sim}153$\,ms to $0.008$\,ms, fast enough for real-time interactive analysis rather than batch processing. An adaptive-$\tau$ regime, where the threshold is calibrated to each query pair's baseline path probability, raised real-data path-found rates from near 0\% to over 80\%. In every case where a path was found, the optimality gap was exactly $0.0000\%$.

Formal correctness was established through five theorems: log-transform order preservation, baseline optimality, convergence at $\tau=0$, conditional optimality, and the all-or-nothing property. All five were confirmed empirically across all 2,160 synthetic and 840 real-data experiment rows (3,000 total).

A scalable and reproducible experimental framework was constructed, incorporating deterministic \texttt{hashlib.md5} seeding, separate timing and memory measurement passes, two graph generators (Erd\H{o}s--R\'{e}nyi and layered), and two real-world datasets (RetailRocket and RecSys~2015 YOOCHOOSE).

Finally, the reachability--speed trade-off was characterized in detail. Increasing $\tau$ yields faster search but reduces path-found rates, from 31.4\% at $\tau=0.001$ to 3.9\% at $\tau=0.5$ on synthetic data. This trade-off is inherent and well understood. On real-world graphs, the pruned algorithm's reachability was initially limited by aggressive fixed thresholding on sparse session-based structures, but the adaptive-$\tau$ approach resolved this limitation.

\section{Limitations}

Several limitations of the current work should be acknowledged. The model operates on a fixed graph snapshot, whereas real e-commerce platforms exhibit time-varying transition probabilities driven by seasonal trends, A/B tests, and user behavior drift. The first-order Markov assumption ($P(\text{next}|\text{current})$) does not capture session-level history effects; for instance, users who arrived via search may behave differently from those who arrived via advertisement. Extending the model to higher-order or variable-length Markov chains would address this limitation at the cost of a larger state space.

Real-data experiments revealed that large item-level graphs yield very low path-found rates under the pruned algorithm, even though the baseline finds paths in all tested configurations. This suggests that selecting source nodes from high-degree hub nodes---as done here for reachability guarantees---combined with aggressive $\tau$ values may not reflect realistic query patterns in production systems. Results may differ for queries originating from low-degree peripheral nodes. Additionally, all experiments were conducted on a single machine using a single thread; the algorithm has not been evaluated in distributed or multi-threaded settings. Finally, the pruning strategy is tied to the log-probability cost model: edges are pruned when $-\log\,p(u,v) > -\log\,\tau$, which requires weights derived from probabilities in $(0,1]$. Applying the same technique to graphs with arbitrary non-negative weights (e.g.\ latency or monetary cost) would require a domain-specific analogue of the threshold criterion.

\section{Future Work}

Several directions merit further investigation. The framework could be extended to support \textbf{dynamic graph updates}, enabling incremental edge weight adjustments as new clickstream data arrives without rebuilding the full graph. Replacing first-order transitions with \textbf{higher-order or variable-length Markov chains} would capture richer user behavior patterns. \textbf{Adaptive $\tau$ selection} methods---such as heuristics or binary search calibrated to a target path-found rate or maximum acceptable runtime---would reduce the need for manual threshold tuning.

Beyond single-objective optimization, a \textbf{multi-objective} formulation incorporating probability, revenue, and user satisfaction through Pareto-optimal path search could yield more actionable business insights. \textbf{Integration with recommendation systems} would allow discovered optimal paths to inform journey-aware recommendations that guide users toward conversion. Finally, adapting the algorithm for \textbf{distributed graph processing frameworks} (e.g., Apache Giraph, GraphX) would enable application to web-scale graphs.
